Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-15 manuscript.tex
label transfer

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
\end{figure}

In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.

@@

\textcolor{red}{TO BE DONE}

>>>>

%\begin{figure}[H]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
%\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
%\end{figure}

In all, these data suggest that \textcolor{red}{SAY SOMETHING HERE}.

@@

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies. We choose cells profiled with Chromium 10X and MARS-seq platforms ((Fig. \ref{Figure_Label_Transfer}A-B), as those were found to be at the extremes of capability to distinguish cell types. After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated both datasets into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph from the resulting unified representation. Note that the original cell type is known for both datasets. We retained cell type annotation for cells derived from 10X, while we assigned a 'Unknown' label to all cells derived from MARS-seq. We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'. Once the matrix has been computed, each cell from the MARS-seq data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they are annotated as different cell types; cells from the MARS-seq should retain their unannotated label if and only if there is not enough evidence to associate them to another group. In parallel, we also relabeled MARS-seq data according to the closest annotated cell in the reference \emph{k}NN graph. Finally we calculated the accuracy of label transfer of both methods using known annotations; we found that the \emph{k}NN approach (Fig. \ref{Figure_Label_Transfer}C) performs worse than an approach based on Stochastic Block Models (Fig. \ref{Figure_Label_Transfer}C). In the latter, approximately 4\% of cells remain unlabeled. Interestingly, we found that for the largest part of these cells, the second choice by affinity was indeed the appropriate one (Fig. \ref{hist_label_second}).

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Label_Transfer.png}
\caption[]{Label transfer using SBM. Annotated data from Chromium 10X (A) and  MARS-seq (B) were merged using Harmony. MARS-seq cells were relabeled according to the \emph{k}NN graph (C) or by exploiting the Stochastic Block Model (D). }\label{Figure_Label_Transfer}
\end{figure}

### 2021-04-15 supplementary.tex
label transfer

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors. Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis). Both the parameters influence the sparseness of the graph.}\label{MPFigure:F56BBDC3-0F0A-4558-8BC2-839E91250DB4}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
\caption[]{Analysis of hematopoietic differentiation. (A) Low dimensional embedding of single cells colored by original cell type and pseudotime. (B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{MPFigure:6C191331-D4E9-4774-A8E0-210B2852D498}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency. Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
\end{figure}

>>>>

%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
%\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors. Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis). Both the parameters influence the sparseness of the graph.}\label{MPFigure:F56BBDC3-0F0A-4558-8BC2-839E91250DB4}
%\end{figure}

%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
%\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
%\end{figure}

%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
%\caption[]{Analysis of hematopoietic differentiation. (A) Low dimensional embedding of single cells colored by original cell type and pseudotime. (B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{MPFigure:6C191331-D4E9-4774-A8E0-210B2852D498}
%\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{hist_label_second.png}
\caption[]{TO BE WRITTEN}\label{hist_label_second}
\end{figure}

%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
%\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency. Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
%\end{figure}
